### 2014-10-08 Studies/2011_San_Francisco/Report/San_Francisco_Fire.tex
San Francisco: Minor edits
San Francisco: Edits based on comments from division reviewer

% PDF metadata
\hypersetup{
  pdftitle={Simulation of a Fire in a Hillside Residential Structure - San Francisco, CA},
  pdfauthor={Overholt, Kristopher J., Weinschenk, Craig G., and Madrzykowski, Daniel},
  pdfsubject={The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on June 2, 2011, within a multi-level, single-family residential structure in San Francisco, CA, that resulted in the death of two firefighters. The inputs for the FDS simulations are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the basement and resulted in ventilation limited (fuel rich) fire conditions in the basement area. The fire originated in the basement, and the interior stairwell acted as a chimney for hot gases in the basement to flow towards areas of lower pressure via the interior stairwell towards the front of the structure. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure areas via the stairwell. The temperature of the gases in the interior stairwell was estimated to be in excess of 700~$^{\circ}$C (1300~$^{\circ}$F), which exceeds the Class~III exposure temperature of 260~$^{\circ}$C (500~$^{\circ}$F). Unfortunately, two E26 firefighters were located in the flow path between the basement and the doors on the front side of the structure. After a Mayday call, both firefighters were removed from the structure and immediate medical treatment was provided. The two firefighters were transported to the local medical center where the lieutenant was pronounced dead and the fire fighter/paramedic died two days later.},
  pdfkeywords={flow path; fire dynamics; LODD; fire modeling; FDS simulation}
}

@@

The fire originated in the basement, and the interior stairwell acted as a chimney for hot gases in the basement to flow towards areas of lower pressure via the interior stairwell towards the front of the structure. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure areas via the stairwell. The temperature of the gases in the interior stairwell was estimated to be in excess of 700~$^{\circ}$C (1300~$^{\circ}$F), which exceeds the Class~III exposure temperature of 260~$^{\circ}$C (500~$^{\circ}$F). Unfortunately, two E26 firefighters were located in the flow path between the basement and the doors on the front side of the structure. After a Mayday call, both firefighters were removed from the structure and immediate medical treatment was provided. The two firefighters were transported to the local medical center where the lieutenant was pronounced dead and the fire fighter/paramedic died two days later.

@@

On June 2, 2011, a fire in a multi-level, wood-frame residential structure claimed the life of two firefighters of the San Francisco Fire Department. NIST examined the fire dynamics of this incident at the request of NIOSH and the San Francisco Fire Department. Computer simulations of the fire incident were conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide} software to provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this multi-level residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the San Francisco Fire Department likely experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.1.2) simulations. The simulations were developed using a combination of knowledge of the fire scenario and appropriate engineering approximations and assumptions. Analysis of the simulation results focuses on the hazardous conditions that developed following the development of a flow path inside of the structure. This document is organized as follows: Section~\ref{sec:incident_summary} provides a summary of the fire incident, Section~\ref{sec:model_description} describes the relevant model input parameters and assumptions that were used to develop the simulations, Section~\ref{sec:model_results} presents the simulation results, and Section~\ref{sec:discussion} discusses the simulation results as they relate to firefighter safety and effectiveness. Appendix~\ref{sec:drawings} contains dimensioned drawings of the basement, first, and second floors of the structure.

@@

The account of events for this incident was documented by the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2011-13~\cite{NIOSH:Bowyer2}. For the purpose of this analysis, the details regarding the timeline are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer2}:
\begin{quote}
On June 02, 2011, a 48 year-old career lieutenant and a 53 year-old fire fighter/paramedic died in a multi-level residential structure fire while searching for the seat of the fire. Note: The residential structure where the fatalities occurred was built on a significantly sloped hillside common throughout the city. The fire floor was one floor below street level. Six companies and three command chiefs were dispatched to a report of an electrical fire at a residential home.

@@

The analysis of the simulation results is focused on the interior conditions on the first floor where the downed firefighters were located and the conditions that resulted in the establishment of a flow path in the interior stairwell leading into the basement. Based on the goals of the analysis and the information collected, several engineering approximations were made. For example, the fire was in the early growth stage when firefighters arrived~\cite{NIOSH:Bowyer2}; however, the exact ignition time, fire growth rate, and heat release rate (HRR) of the fire were not known but were estimated using documentation of the potential fuel load within the structure.

@@

The structure involved in this fire incident was a four-story residential building with a flat roof that consisted of a bitumen membrane. The exterior was primarily stucco with large windows on the top three stories of the rear side of the structure. The structure had exterior dimensions of 8.53~m (28~ft) by 17.73~m (58~ft) and a height of 5.9~m (19~ft) above street level, as shown in Fig.~\ref{fig:floor_plan}. The right side of the structure shared a common fire wall with another residential structure.

@@

The fire, which is described in Section~\ref{sec:fires}, originated in the finished living room on the basement level (near the rear side of the structure), which had a floor area of 72~m$^2$ (775~ft$^2$). More detailed discussion on the impact of the fire on the interior conditions is provided in Section~\ref{sec:model_results}. Figure~\ref{fig:post_exterior} shows the exterior of the front and rear sides of the structure after the incident. Note that the majority of the exterior damage occurred above the openings located on the basement floor (where the wooden balcony was located). Fully dimensioned drawings of the second floor, first floor, and basement are shown in Figs.~\ref{fig:drawing_second_floor}, \ref{fig:drawing_first_floor} and \ref{fig:drawing_basement}, respectively.

@@

\begin{figure}[!ht]
\includegraphics[width=5.0in]{../Figures/Plan_Second_Floor}
\includegraphics[width=5.0in]{../Figures/Plan_First_Floor}
\includegraphics[width=5.0in]{../Figures/Plan_Basement_Floor}
\caption[Plan view of second floor, first floor, and basement floor of structure.]
{Plan view of second floor (top), first floor (middle), and basement floor (bottom) of structure. The fire originated in the living room on the basement floor. Stairs and rooms are identified using information collected by NIST.}
\label{fig:floor_plan}
\end{figure}

@@

The basement primarily contained upholstered polyurethane furniture fuel items, and the balcony was constructed of wood. The HRR of the initial couch fire was based on an experimentally measured HRR for a mattress fire~\cite{madrzykowski2009fire} to represent the fire growth rate of a similar household furniture item. The HRRs of the additional furniture items were estimated based on previous experimental work that characterized the HRR of similar upholstered furniture fuel items~\cite{madrzykowski2009fire,Madrzykowski:2,Babrauskas:1}. Other combustible items and furnishings such as tables, stools, bookshelves, and carpet were present in the basement living room. However, to simplify the analysis and specification of the design fire, only the large upholstered furniture fuel items and balcony were included in the simulation because they were considered to have contributed the largest amount of energy in this fire scenario.

@@

Polyurethane fuel was specified for the interior basement fuel items, which can be represented by the chemical formula, $\C\H_{1.8}\O_{0.30}\N_{0.05}$, with specified product yields of soot ($y_{\mathrm{s}}=0.131 \; \mathrm{kg}/\mathrm{kg}$) and carbon monoxide ($y_{\mathrm{CO}}=0.010 \; \mathrm{kg}/\mathrm{kg}$)~\cite{SFPE:Tewarson}. The product yields are expressed in terms of the amount soot or carbon monoxide emitted per unit mass of fuel consumed~(kg/kg) and can be found in the Society of Fire Protection Engineers (SFPE) Handbook~\cite{SFPE:Tewarson}. A balanced chemical reaction for polyurethane combustion can be written as:

\begin{multline}
\C\H_{1.8}\O_{0.30}\N_{0.05} + 5.22(0.208\,\O_{2} + 0.783\,\N_{2} + 8.34 \times 10^{-3}\,\H_{2}\O + 3.87 \times 10^{-4}\,\C\O_{2}) \\
\rightarrow 6.06(0.129\,\C\O_{2} + 0.156\,\H_{2}\O + 0.001\,\C\O + 0.035\,\C + 0.679\,\N_{2})
\label{eq:fuel_polyurethane}
\end{multline}

@@

Wood fuel was specified for the exterior basement balcony fuel items, which can be represented by the chemical formula, $\C\H_{1.7}\O_{0.74}\N_{0.002}$, with specified product yields of soot ($y_{\mathrm{s}}=0.015 \; \mathrm{kg}/\mathrm{kg}$) and carbon monoxide ($y_{\mathrm{CO}}=0.004 \; \mathrm{kg}/\mathrm{kg}$)~\cite{SFPE:Tewarson}. A balanced chemical reaction for wood combustion can be written as:

\begin{multline}
\C\H_{1.7}\O_{0.74}\N_{0.002} + 4.91(0.208\,\O_{2} + 0.783\,\N_{2} + 8.34 \times 10^{-3}\,\H_{2}\O + 3.87 \times 10^{-4}\,\C\O_{2}) \\
\rightarrow 5.74(0.168\,\C\O_{2} + 0.155\,\H_{2}\O + 6.37 \times 10^{-4}\,\C\O + 5.57 \times 10^{-3}\,\C + 0.670\,\N_{2})
\label{eq:fuel_wood}
\end{multline}

The heat of combustion of polyurethane and wood were specified as 26,200~kJ/kg and 16,400~kJ/kg, respectively, based on data provided in the SFPE Handbook~\cite{SFPE:Tewarson}. The heat of combustion represents the amount of energy released per unit mass of the fuel. The following text defines Eqs.~\ref{eq:fuel_polyurethane} and \ref{eq:fuel_wood} for an FDS input file with the two fuels and reaction properties discussed above:

@@

A time-dependent burning rate was specified for each of the fuel packages in terms of a fuel mass flux. The fuel mass flux is the amount of fuel vapor per unit area that is released from the surface of each fuel package shown in Fig.~\ref{fig:fuel_placement}. The specified time-dependent fuel mass flux for each fuel package corresponds to the HRRs that were described earlier. The HRR from each individual fuel package contributes to the total HRR shown in Fig.~\ref{fig:hrr}. FDS then determines the amount of combustion that occurs throughout the simulation based on the amount of available fuel and oxygen at a given location. The following lines define the fire surfaces for each fuel package in the FDS input file:

\begin{lstlisting}
&SURF ID='COUCH_RIGHT_INITIAL',
    SPEC_ID='POLYURETHANE',
    MASS_FLUX = 5.271815E-6,
    RAMP_MF='COUCH_RIGHT_INITIAL_RAMP' /

&SURF ID='COUCH_MIDDLE_FLASHOVER',
    SPEC_ID='POLYURETHANE',
    MASS_FLUX = 1.037172E-5,
    RAMP_MF='COUCH_MIDDLE_FLASHOVER_RAMP' /

&SURF ID='COUCH_LEFT_FLASHOVER',
    SPEC_ID='POLYURETHANE',
    MASS_FLUX = 5.271815E-6,
    RAMP_MF='COUCH_LEFT_FLASHOVER_RAMP' /

&SURF ID='LOUNGE_FLASHOVER',
    SPEC_ID='POLYURETHANE',
    MASS_FLUX = 7.853485E-6,
    RAMP_MF='LOUNGE_FLASHOVER_RAMP' /

&SURF ID='CHAIR_FLASHOVER',
    SPEC_ID='POLYURETHANE',
    MASS_FLUX = 7.918660E-6,
    RAMP_MF='CHAIR_FLASHOVER_RAMP' /

&SURF ID='WOOD_EXTERIOR',
    SPEC_ID='WOOD',
    MASS_FLUX = 0.003049,
    RAMP_MF='WOOD_EXTERIOR_RAMP' /
\end{lstlisting}

Note that the mass flux values here are unique for each fuel item and are specific to the input file used for these simulations. The key point is that each fuel item was modeled as a 3D object with surfaces from which a specified amount of fuel flows into the domain, mixes with air, and burns.

@@

While the fires were represented as fuel sources with a constant area (Section~\ref{sec:fires}), from a heat transfer perspective, it is important to define the material properties of the ceiling, walls, and floors (density, thermal conductivity, specific heat, and thickness) to account for heat transfer and energy storage. In this study, the material properties of gypsum board~\cite{WAKILI2007} were specified on the finished walls and ceilings of the structure. The following lines define the gypsum material and surface properties in the FDS input file:

\begin{lstlisting}
&MATL ID          = 'GYPSUM'
    FYI           = 'Wakili - Journal of Fire Sciences 2007' 
    CONDUCTIVITY  = 0.28
    SPECIFIC_HEAT = 1.00
    DENSITY       = 810. /

&SURF ID      = 'WALL'
    DEFAULT   = .TRUE.
    COLOR     = 'DARK SEA GREEN 4'
    MATL_ID   = 'GYPSUM'
    THICKNESS = 0.012 /
\end{lstlisting}

@@

For the simulation, a measure of how well the flow field is resolved can be estimated by using the non-dimensional expression $D^*/\dx$. Here, $D^*$ is the characteristic fire diameter, $\dx$ is the nominal size of a mesh cell, and $\dQ$ is the total heat release rate of the fire:
\begin{equation}
D^* = \left(
     \frac{\dQ}{\rho_\infty \, c_p \, T_\infty \, \sqrt{g}}
     \right)^\frac{2}{5} 
\label{eq:mesh}
\end{equation}
From the FDS User's Guide~\cite{FDS_Users_Guide}, the characteristic fire diameter is related to the characteristic fire size via the relation $Q^* = (D^*/D)^{5/2}$. Here, $D$ is the physical diameter of the base of the fire specified in the simulation. Based on validation work performed for the U.S. Nuclear Regulatory Commission, $D^*/\dx$ values that ranged between 4 and 16 produced results that were adequate for engineering calculations~\cite{NUREG_1824}. Following Eq.~\ref{eq:mesh} and using a grid cell size of 10~cm, the ratio of the characteristic fire diameter to cell size ($D^*/\dx$) was 18 for the initial couch fire (combined HRR of 5~MW), 31 for the secondary flashover fire in the basement (combined HRR of 19~MW), and 32 for the maximum prescribed fire size in the structure (combined HRR of 20~MW). Therefore, the grid resolution used in this model results in a $D^*/\dx$ ratio that exceeds typical engineering values.

As discussed in Section~\ref{sec:geometry}, the structure had exterior dimensions of 8.53~m (28~ft) by 17.73~m (58~ft) and a height of 5.9~m (19~ft) above street level. To ensure adequately resolved fluid flow in and out of the structure, the computational domain was extended beyond the volume of the structure. The dimensions of the computational domain were 11.5~m (38~ft) by 22~m (72~ft) and a height of 13.6~m (45~ft). A grid resolution of 10~cm (3.9~in) was used, which resulted in a total of approximately 3.4 million computational cells. As a result, all of the input model geometry lengths, ventilation openings (holes, doors, and windows), and fire areas snapped to the nearest 10~cm. Figure~\ref{fig:SMV_exterior_grid} shows the front and rear sides of the structure rendered in Smokeview with the 10~cm mesh.

@@

In an FDS simulation, a prescribed HRR is specified, which results in a specified amount of fuel vapor (or pyrolyzate) being released from a fuel surface. FDS then determines the amount of combustion that occurs throughout the simulation based on the amount of available fuel and oxygen at a given location. As a result, the prescribed HRR that was input into the simulation can be different than the HRR predicted by the simulation based on the ventilation conditions.

Figure~\ref{fig:hrr_w_fds} shows the HRR vs. time based on the fires that were described in Section~\ref{sec:fires}. In this figure, the solid line represents the specified (or prescribed) HRR that was input into the simulation, and the dashed line represents the actual (or predicted) HRR that was output from the simulation based on the ventilation conditions. The vertical dashed line represents the time at which the first basement window on the rear side of the structure failed (342~s). Additional windows on the rear side of the basement continued to fail within the next two minutes.

@@

Gases will always flow from a region of high pressure towards a region of lower pressure. Once the rear basement windows failed, the gases in the basement at an elevated temperature and pressure flowed into the interior stairwell towards the lower pressure regions in the house and exited the structure via the garage door and front door. The velocity contours shown in Fig.~\ref{fig:smv_velocity} indicate the magnitude of flow velocities within the structure before and after the rear basement window failures.

@@

After the rear basement windows failed, a flow path was established between the basement living room area and the doors located on the front side of the structure (the front door and the garage door). The rear basement window failures resulted in a rapid change in the conditions within the flow path. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure areas via the interior stairwell. The two E26 firefighters were located in the flow path between the basement and the doors on the front side of the structure.

@@

After the primary search was conducted, closing the interior doors (i.e., door control) could have inhibited the establishment of a flow path in the interior stairwell. A 360$^\circ$ scene size-up by the arriving firefighters could have helped determine the location of the fire in the basement and identify potential flow paths within the structure. The interior stairwell acted as a chimney for hot gases in the basement to flow towards areas of lower pressure and vent openings at the front of the structure. Firefighters should avoid placing themselves within a flow path where elevated temperatures and flow velocities can present hazardous conditions and increase the rate of heat transfer to firefighting gear via convection.

@@

Unfortunately, two E26 firefighters were located in the flow path between the basement and the doors on the front side of the structure. After a Mayday call, both firefighters were removed from the structure and immediate medical treatment was provided. The two firefighters were transported to the local medical center where the lieutenant was pronounced dead and the fire fighter/paramedic died two days later.

@@

\begin{figure}[!ht]
\includegraphics[width=6.5in]{../Figures/Drawing_Second_Floor_Metric}
\caption[Dimensioned drawing of the second floor.]
{Dimensioned drawing of the second floor. The measurements are accurate to within 15~cm (6~in).}
\label{fig:drawing_second_floor}
\end{figure}

\begin{figure}[!ht]
\includegraphics[width=6.5in]{../Figures/Drawing_First_Floor_Metric}
\caption[Dimensioned drawing of the first floor.]
{Dimensioned drawing of the first floor. The measurements are accurate to within 15~cm (6~in).}
\label{fig:drawing_first_floor}
\end{figure}

\begin{figure}[!ht]
\includegraphics[width=6.5in]{../Figures/Drawing_Basement_Metric}
\caption[Dimensioned drawing of the basement floor.]
{Dimensioned drawing of the basement floor. The measurements are accurate to within 15~cm (6~in).}
\label{fig:drawing_basement}
\end{figure}

>>>>

% PDF metadata
\hypersetup{
  pdftitle={Simulation of a Fire in a Hillside Residential Structure - San Francisco, CA},
  pdfauthor={Overholt, Kristopher J., Weinschenk, Craig G., and Madrzykowski, Daniel},
  pdfsubject={The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on June 2, 2011, within a multi-level, single-family residential structure in San Francisco, CA, that resulted in the death of two firefighters. The inputs for the FDS simulations are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the basement and resulted in ventilation limited (fuel rich) fire conditions in the basement area. The fire originated in the basement, and the interior stairwell acted as a chimney for hot gases in the basement to flow towards areas of lower pressure via the interior stairwell towards the front of the structure. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure areas via the stairwell. The temperature of the gases in the interior stairwell was estimated to be in excess of 700~$^{\circ}$C (1300~$^{\circ}$F). Two firefighters were located in the flow path between the basement and the doors on the front side of the structure, were exposed to these elevated temperatures, and later died as a result of their injuries.},
  pdfkeywords={flow path; fire dynamics; LODD; fire modeling; FDS simulation}
}

@@

The fire originated in the basement, and the interior stairwell acted as a chimney for hot gases in the basement to flow towards areas of lower pressure via the interior stairwell towards the front of the structure. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure areas via the stairwell. The temperature of the gases in the interior stairwell was estimated to be in excess of 700~$^{\circ}$C (1300~$^{\circ}$F). Two firefighters were located in the flow path between the basement and the doors on the front side of the structure, were exposed to these elevated temperatures, and later died as a result of their injuries.

@@

On June 2, 2011, a fire in a multi-level, wood-frame residential structure claimed the life of two firefighters of the San Francisco Fire Department. NIST examined the fire dynamics of this incident at the request of the National Institute for Occupational Safety and Health (NIOSH) and the San Francisco Fire Department. Computer simulations of the fire incident were conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide} software to provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of fire-induced flow paths (including temperature, pressure, and fire conditions) in this multi-level residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the San Francisco Fire Department likely experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.1.2) simulations. The simulations were developed using a combination of knowledge of the fire scenario and appropriate engineering approximations and assumptions. Analysis of the simulation results focuses on the hazardous conditions that developed following the development of a flow path inside of the structure. This document is organized as follows: Section~\ref{sec:incident_summary} provides a summary of the fire incident, Section~\ref{sec:model_description} describes the relevant model input parameters and assumptions that were used to develop the simulations, Section~\ref{sec:model_results} presents the simulation results, and Section~\ref{sec:discussion} discusses the simulation results as they relate to firefighter safety and effectiveness. Appendix~\ref{sec:drawings} contains dimensioned drawings of the basement, first, and second floors of the structure.

@@

The account of events for this incident was documented by NIOSH Fire Fighter Fatality Investigation and Prevention Program Report \#F2011-13~\cite{NIOSH:Bowyer2}. For the purpose of this analysis, the details regarding the timeline are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer2}:
\begin{quote}
On June 02, 2011, a 48 year-old career lieutenant and a 53 year-old fire fighter/paramedic died in a multi-level residential structure fire while searching for the seat of the fire. Note: The residential structure where the fatalities occurred was built on a significantly sloped hillside common throughout the city. The fire floor was one floor below street level. Six companies and three command chiefs were dispatched to a report of an electrical fire at a residential home.

@@

The analysis of the simulation results is focused on the interior conditions on the first floor where the downed firefighters were located and the conditions that resulted in the establishment of a flow path in the interior stairwell leading into the basement. Based on the goals of the analysis and the information collected, several assumptions were made. For example, the fire was in the early growth stage when firefighters arrived~\cite{NIOSH:Bowyer2}; however, the exact ignition time, fire growth rate, and heat release rate (HRR) of the fire were not known but were estimated using documentation of the potential fuel load within the structure.

@@

The structure involved in this fire incident was a four-story residential building with a flat roof. The exterior was primarily stucco with large windows on the top three stories of the rear side of the structure. The structure had exterior dimensions of 8.5~m (28~ft) by 17.7~m (58~ft) and a height of 5.9~m (19~ft) above street level, as shown in Fig.~\ref{fig:floor_plan}. The right side of the structure shared a common fire wall with another residential structure.

@@

The fire, which is described in Section~\ref{sec:fires}, originated in the finished living room on the basement level (near the rear side of the structure), which had a floor area of 72~m$^2$ (775~ft$^2$). More detailed discussion on the impact of the fire on the interior conditions is provided in Section~\ref{sec:model_results}. Figure~\ref{fig:post_exterior} shows the exterior of the front and rear sides of the structure after the incident. Note that the majority of the exterior damage occurred above the openings located on the basement floor (where the wooden balcony was located). Fully dimensioned drawings of the second floor, first floor, and basement are shown in Appendix~\ref{sec:drawings}.

@@

\begin{figure}[!ht]
\includegraphics[width=5.0in]{../Figures/Plan_Second_Floor}
\includegraphics[width=5.0in]{../Figures/Plan_First_Floor}
\includegraphics[width=5.0in]{../Figures/Plan_Basement_Floor}
\caption[Plan view of second floor, first floor, and basement floor of structure.]
{Plan view of second floor (top), first floor (middle), and basement floor (bottom) of structure. The fire originated in the living room on the basement floor. Stairs and rooms are identified using information collected by NIST. The sub-basement level was not connected to the other three levels of the house and is not shown.}
\label{fig:floor_plan}
\end{figure}

@@

The basement primarily contained large upholstered polyurethane furniture items, and the balcony was constructed of wood. The HRR of the initial couch fire was based on an experimentally measured HRR for a mattress fire~\cite{madrzykowski2009fire} to represent the fire growth rate of a similar household furniture item. The HRRs of the additional furniture items were estimated based on previous experimental work that characterized the HRR of similar upholstered furniture fuel items~\cite{madrzykowski2009fire,Madrzykowski:2,Babrauskas:1}. Other combustible items and furnishings such as tables, stools, bookshelves, and carpet were present in the basement living room. However, to simplify the analysis and specification of the design fire, only the large upholstered furniture fuel items and balcony were included in the simulation because they were considered to have contributed the largest amount of energy in this fire scenario.

@@

Polyurethane fuel was specified for the interior basement fuel items, which can be represented by the chemical formula, $\C\H_{1.8}\O_{0.30}\N_{0.05}$, with specified product yields of soot ($y_{\mathrm{s}}=0.131 \; \mathrm{kg}/\mathrm{kg}$) and carbon monoxide ($y_{\mathrm{CO}}=0.010 \; \mathrm{kg}/\mathrm{kg}$)~\cite{SFPE:Tewarson}. The product yields are expressed in terms of the amount soot or carbon monoxide emitted per unit mass of fuel consumed~(kg/kg) and can be found in the Society of Fire Protection Engineers (SFPE) Handbook~\cite{SFPE:Tewarson}. A balanced chemical reaction for polyurethane combustion can be written as:
\begin{multline}
\C\H_{1.8}\O_{0.30}\N_{0.05} + 5.22(0.208\,\O_{2} + 0.783\,\N_{2} + 8.34 \times 10^{-3}\,\H_{2}\O + 3.87 \times 10^{-4}\,\C\O_{2}) \\
\rightarrow 6.06(0.129\,\C\O_{2} + 0.156\,\H_{2}\O + 0.001\,\C\O + 0.035\,\C + 0.679\,\N_{2})
\label{eq:fuel_polyurethane}
\end{multline}

@@

Wood fuel was specified for the exterior basement balcony fuel items, which can be represented by the chemical formula, $\C\H_{1.7}\O_{0.74}\N_{0.002}$, with specified product yields of soot ($y_{\mathrm{s}}=0.015 \; \mathrm{kg}/\mathrm{kg}$) and carbon monoxide ($y_{\mathrm{CO}}=0.004 \; \mathrm{kg}/\mathrm{kg}$)~\cite{SFPE:Tewarson}. A balanced chemical reaction for wood combustion can be written as:
\begin{multline}
\C\H_{1.7}\O_{0.74}\N_{0.002} + 4.91(0.208\,\O_{2} + 0.783\,\N_{2} + 8.34 \times 10^{-3}\,\H_{2}\O + 3.87 \times 10^{-4}\,\C\O_{2}) \\
\rightarrow 5.74(0.168\,\C\O_{2} + 0.155\,\H_{2}\O + 6.37 \times 10^{-4}\,\C\O + 5.57 \times 10^{-3}\,\C + 0.670\,\N_{2})
\label{eq:fuel_wood}
\end{multline}
The heat of combustion of polyurethane and wood were specified as 26,200~kJ/kg and 16,400~kJ/kg, respectively, based on data provided in the SFPE Handbook~\cite{SFPE:Tewarson}. The heat of combustion represents the amount of energy released per unit mass of the fuel. The following text defines Eqs.~\ref{eq:fuel_polyurethane} and \ref{eq:fuel_wood} for an FDS input file with the two fuels and reaction properties discussed above:

@@

A time-dependent burning rate was specified for each of the fuel packages in terms of a fuel mass flux. The fuel mass flux is the amount of fuel vapor per unit area that is released from the surface of each fuel package shown in Fig.~\ref{fig:fuel_placement}. The specified time-dependent fuel mass flux for each fuel package corresponds to the HRRs that were described earlier. The HRR from each individual fuel package contributes to the total HRR shown in Fig.~\ref{fig:hrr}. FDS then determines the amount of combustion that occurs throughout the simulation based on the amount of available fuel and oxygen at a given location. Each fuel item was modeled as a 3D object with surfaces from which a specified amount of fuel flows into the domain, mixes with air, and burns.

While the fires were represented as fuel sources with a constant area (Section~\ref{sec:fires}), from a heat transfer perspective, it is important to define the material properties of the ceiling, walls, and floors (density, thermal conductivity, specific heat, and thickness) to account for heat transfer and energy storage. In this study, the material properties of gypsum board~\cite{WAKILI2007} were specified on the finished walls and ceilings of the structure. Table~\ref{tab:model_parameters} shows the material properties of the gypsum board that were used in the simulation.

@@

@@

For the simulation, a measure of how well the flow field is resolved can be estimated by using the non-dimensional expression $D^*/\dx$. Here, $D^*$ is the characteristic fire diameter, $\dx$ is the nominal size of a mesh cell, and $\dQ$ is the total heat release rate of the fire:
\begin{equation}
D^* = \left(
     \frac{\dQ}{\rho_\infty \, c_p \, T_\infty \, \sqrt{g}}
     \right)^\frac{2}{5} 
\label{eq:mesh}
\end{equation}
From the FDS User's Guide~\cite{FDS_Users_Guide}, the characteristic fire diameter is related to the characteristic fire size via the relation $Q^* = (D^*/D)^{5/2}$. Here, $D$ is the physical diameter of the base of the fire specified in the simulation. Based on validation work performed for the U.S. Nuclear Regulatory Commission, $D^*/\dx$ values that ranged between 4 and 16 produced results that were adequate for engineering calculations~\cite{NUREG_1824}. Following Eq.~\ref{eq:mesh} and using a grid cell size of 10~cm, the ratio of the characteristic fire diameter to cell size ($D^*/\dx$) was 18 for the initial couch fire (combined HRR of 5~MW), 31 for the secondary flashover fire in the basement (combined HRR of 19~MW), and 32 for the maximum prescribed fire size in the structure (combined HRR of 20~MW). Therefore, the grid resolution used in this model results in a $D^*/\dx$ ratio that exceeds typical engineering values. Note that Eq.~\ref{eq:mesh} relates to gas-phase phenomena, which is the primary concern of this study because the fires were explicitly specified instead of attempting to predict the amount of pyrolysis or flame spread in the scenario.

As discussed in Section~\ref{sec:geometry}, the structure had exterior dimensions of 8.5~m (28~ft) by 17.7~m (58~ft) and a height of 5.9~m (19~ft) above street level. To ensure adequately resolved fluid flow in and out of the structure, the computational domain was extended beyond the volume of the structure. The dimensions of the computational domain were 11.5~m (38~ft) by 22~m (72~ft) and a height of 13.6~m (45~ft). A grid resolution of 10~cm (3.9~in) was used, which resulted in a total of approximately 3.4 million computational cells. As a result, all of the input model geometry lengths, ventilation openings (holes, doors, and windows), and fire areas snapped to the nearest 10~cm. Figure~\ref{fig:SMV_exterior_grid} shows the front and rear sides of the structure rendered in Smokeview with the 10~cm mesh.

@@

In an FDS simulation, a HRR is specified, which results in a specified amount of fuel vapor (or pyrolyzate) being released from a fuel surface. FDS then determines the amount of combustion that occurs throughout the simulation based on the amount of available fuel and oxygen at a given location. As a result, the prescribed HRR that was input into the simulation can be different than the HRR predicted by the simulation based on the ventilation conditions.

Figure~\ref{fig:hrr_w_fds} shows the HRR vs. time based on the fires that were described in Section~\ref{sec:fires}. In this figure, the solid line represents the specified (or prescribed) HRR that was input into the simulation, and the dashed line represents the predicted HRR that was output from the simulation based on the ventilation conditions. The vertical dashed line represents the time at which the first basement window on the rear side of the structure failed (342~s). Additional windows on the rear side of the basement continued to fail within the next two minutes.

@@

Gases will always flow from a region of high pressure towards a region of lower pressure. Once the rear basement windows failed, the gases in the basement at an elevated temperature and pressure flowed upward into the interior stairwell and exited the structure via the garage door and front door. The velocity contours shown in Fig.~\ref{fig:smv_velocity} indicate the magnitude of flow velocities within the structure before and after the rear basement window failures.

@@

After the rear basement windows failed, a flow path was established between the basement living room area and the doors located on the front side of the structure (the front door and the garage door). The rear basement window failures resulted in a rapid change in the conditions within the flow path. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure areas via the interior stairwell. The two firefighters were located in the flow path between the basement and the doors on the front side of the structure.

@@

A 360$^\circ$ scene size-up by arriving firefighters can help determine the location of the fire and identify potential flow paths within a structure. The interior stairwell acted as a chimney for hot gases in the basement to flow towards areas of lower pressure and vent openings at the front of the structure. Firefighters should avoid placing themselves within a flow path where elevated temperatures and flow velocities can present hazardous conditions and increase the rate of heat transfer to firefighting gear via convection. Door control can also be used to avoid creating inlet and outlet vents that could result in the establishment of a flow path.

@@

Two firefighters were located in the flow path between the basement and the doors on the front side of the structure. After a call for assistance, both firefighters were removed from the structure and immediate medical treatment was provided. The two firefighters were transported to the local medical center where the lieutenant was pronounced dead and the fire fighter/paramedic died two days later.

@@

\begin{sidewaysfigure}[!ht]
\includegraphics[width=8in]{../Figures/Drawing_Second_Floor_Metric}
\caption[Dimensioned drawing of the second floor.]
{Dimensioned drawing of the second floor. The measurements are accurate to within 15~cm (6~in).}
\label{fig:drawing_second_floor}
\end{sidewaysfigure}

\begin{sidewaysfigure}[!ht]
\includegraphics[width=8in]{../Figures/Drawing_First_Floor_Metric}
\caption[Dimensioned drawing of the first floor.]
{Dimensioned drawing of the first floor. The measurements are accurate to within 15~cm (6~in).}
\label{fig:drawing_first_floor}
\end{sidewaysfigure}

\begin{sidewaysfigure}[!ht]
\includegraphics[width=8in]{../Figures/Drawing_Basement_Metric}
\caption[Dimensioned drawing of the basement floor.]
{Dimensioned drawing of the basement floor. The measurements are accurate to within 15~cm (6~in).}
\label{fig:drawing_basement}
\end{sidewaysfigure}
